---
title: "Geography details for Problem Set 3"
author: "Carole Voulgaris"
format: html
editor: visual
execute: 
  warning: false
  message: false
---

## Load the packages

First load the packages:

```{r}

options(java.parameters = '-Xmx2G')  

library(tidyverse) 
library(sf) 
library(r5r) 
library(here)
library(knitr)
library(tidytransit)

```

## Initialize R5

Now initialize an R5 core for each of these study areas:

-   Boise City, Idaho

-   Santa Fe, New Mexico

```{r}
boise_core <- here("network-files",
                "Boise-City-ID") |>
  setup_r5()

santa_fe_core <- here("network-files",
                "Santa-Fe-NM") |>
  setup_r5()
```

## Map streets and routes

We can generate line layers of the streets and routes that are included in our network.

```{r}
boise_streets <- street_network_to_sf(boise_core)$edges
boise_routes <- transit_network_to_sf(boise_core)$routes

santa_fe_streets <- street_network_to_sf(santa_fe_core)$edges
santa_fe_routes <- transit_network_to_sf(santa_fe_core)$routes

```

And we can plot them on a map.

### Boise

Here's Boise:

```{r}
ggplot(boise_streets) +
  geom_sf(color = "gray") +
  geom_sf(data = boise_routes,
          color = "red") +
  theme_void()
```

The transit network is mostly limited to Boise City, but the street network is for a rectangle containing the entire metro area. Let's define our study area as just the area within the metro area (or a box containing the metro area) that is served by transit.

We'll draw a box around the transit service area:

```{r}
boise_routes_bbox <- st_bbox(boise_routes) |>
  st_as_sfc()

```

And then we'll draw a box around the street network:

```{r}
boise_streets_bbox <- st_bbox(boise_streets) |>
  st_as_sfc()
```

And now we'll define a box that's just the overlapping area between those two boxes.

```{r}
boise_area <- st_intersection(boise_routes_bbox, 
                              boise_streets_bbox)
```

Here's a map showing those three boxes. In this case, the box for the study area is the same as the box for the transit routes.

```{r}
ggplot() +
  geom_sf(data = boise_streets,
          color = "gray") +
  geom_sf(data = boise_routes,
          color = "red") +
  geom_sf(data = boise_streets_bbox,
          color = "magenta",
          fill = NA) +
  geom_sf(data = boise_routes_bbox,
          color = "blue",
          linewidth = 2, 
          fill = NA) +
  geom_sf(data = boise_area,
          color = "green",
          fill = NA) +
  theme_void()
```

We'll filter our transit network and street network to only include the area in that overlapping area. While we're add it, we'll convert it to a "MULTILINESTRING" datatype, which will facilitate exporting these layers to a format that we can work with in ArcGIS software if we want to later.

```{r}
boise_area_routes <- st_intersection(boise_routes, boise_area) |>
  st_cast("MULTILINESTRING")

boise_area_streets <- st_intersection(boise_streets, boise_area) |>
  st_cast("MULTILINESTRING")
```

And now we can plot just the roads and streets in our study area:

```{r}
ggplot() +
  geom_sf(data = boise_area_streets,
          color = "gray") +
  geom_sf(data = boise_area_routes,
          color = "red") +
  theme_void()
```

### Santa Fe

Now let's take a look at our routes and streets in Santa Fe:

```{r}
ggplot() +
  geom_sf(data = santa_fe_streets,
          color = "gray") +
  geom_sf(data = santa_fe_routes,
          color = "red") +
  theme_void()
```

This time, not only have a road network that extends beyond our transit network, we also have a transit network that extends beyond the Santa Fe metro area (to the Albuquerque metro area).

Again, we'll define our study area as just the area within the metro area (or a box containing the metro area) that is served by transit.

We'll draw a box around the transit service area:

```{r}
santa_fe_routes_bbox <- st_bbox(santa_fe_routes) |>   
  st_as_sfc() 
```

And then we'll draw a box around the street network:

```{r}
santa_fe_streets_bbox <- st_bbox(santa_fe_streets) |>   
  st_as_sfc()
```

And now we'll define a box that's just the overlapping area between those two boxes.

```{r}
santa_fe_area <- st_intersection(santa_fe_routes_bbox,                                
                                 santa_fe_streets_bbox)
```

Here's a map showing those three boxes.

```{r}
ggplot() +
  geom_sf(data = santa_fe_streets,
          color = "gray") +
  geom_sf(data = santa_fe_routes,
          color = "red") +
  geom_sf(data = santa_fe_streets_bbox,
          color = "magenta",
          linewidth = 2,
          fill = NA) +
  geom_sf(data = santa_fe_routes_bbox,
          color = "blue",
          linewidth = 2, 
          fill = NA) +
  geom_sf(data = santa_fe_area,
          color = "green",
          fill = NA) +
  theme_void()
```

Again, we'll filter our transit network and street network to only include the area in that overlapping area (and convert it to a "MULTILINESTRING" datatype).

```{r}
santa_fe_area_routes <- st_intersection(santa_fe_routes, santa_fe_area) |>
  st_cast("MULTILINESTRING")

santa_fe_area_streets <- st_intersection(santa_fe_streets, santa_fe_area) |>
  st_cast("MULTILINESTRING")
```

And we can plot that:

```{r}
ggplot() +
  geom_sf(data = santa_fe_area_streets,
          color = "gray") +
  geom_sf(data = santa_fe_area_routes,
          color = "red") +
  theme_void()
```

But here, because of the orientation of the transit service to Albuquerque, there's still a lot of area in that overlapping box that isn't served by transit at all.

We could narrow the study area further, by creating a new (smaller) bounding box around the newly filtered transit network.

```{r}
santa_fe_transit_area <- st_bbox(santa_fe_area_routes) |>
  st_as_sfc()
```

Now here is that new bounding box on a map:

```{r}
ggplot() +
  geom_sf(data = santa_fe_area_streets,
          color = "gray") +
  geom_sf(data = santa_fe_area_routes,
          color = "red") +
  geom_sf(data = santa_fe_transit_area,
          color = "blue",
          fill = NA) +
  theme_void()
```

And let's filter the streets to include only that smaller area:

```{r}
santa_fe_area_streets <- santa_fe_area_streets |>
  st_filter(santa_fe_transit_area)
  
```

And here are the routes and streets in our new study area:

```{r}
ggplot() +
  geom_sf(data = santa_fe_area_streets,
          color = "gray") +
  geom_sf(data = santa_fe_area_routes,
          color = "red") +
  theme_void()
```

##  Stop R5

Remember to stop R5 when you're done.

```{r}
stop_r5()
```
